Parsing Postfix log files

### 2008-11-22 doc/logparser-implementation.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

    \item The log files are whole and complete: nothing has been removed,
        either deliberately or accidentally (e.g.\ log rotation gone awry,
        file system filling up, logging system unable to cope with the
        volume of log messages).  On a well run system it is extremely
        unlikely that any of these problems will arise, though it is of
        course possible, particularly when suffering a deluge of spam or a
        mail loop.

@@

%    \item [cluster\_group] A reference to the \texttt{cluster\_group}
%        table.  That table is used by the Decision Tree algorithm described
%        in a separate document.

@@

    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
        local address when receiving mail, the remote address when sending
        mail.

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the \IP{} address could not be resolved to a
        hostname via DNS\@.

    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
        address when receiving mail, the local address when sending mail.

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the \IP{} address could not be resolved to a
        hostname via DNS\@.

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \SMTP{} command, but it
        is quite simple to rectify this, as described in
        \sectionref{logging helo}.

@@

    \item [start] The timestamp of the first log line, in seconds since the
        epoch (explained in the glossary, \sectionref{Glossary}).

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \SMTP{} command --- a
        mechanism that is quite useful for testing new restrictions.  This
        field will be 1 if the log line parsed was a warning rather than a
        real rejection, or 0 for a real rejection.  This field should be
        ignored if the result is not a rejection, i.e.\ the action field of
        the associated rule is not \texttt{REJECTION}.

    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
        general an \SMTP{} code is only present in a log line for a
        rejection or final delivery; results initially missing an \SMTP{}
        code will duplicate the \SMTP{} code of other results in that
        connection.  Some final delivery log lines do not contain an
        \SMTP{} code: in those cases the code is specified by the rule,
        based on the success or failure represented by the log line.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from a \DNSBL{}\@.

    \item [timestamp] The time at which the log line was logged, in seconds
        since the epoch.

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \DNSBL{}
services, policy servers, or custom rejection messages.  To facilitate
parsing new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions can be difficult, but adding new rules
to parse new rejection messages is trivial, and occurs much more
frequently.  A program is provided to ease the process of creating new
rules from unparsed log lines, based on the algorithm developed by Risto
Vaarandi for his \SLCT{}~\cite{slct-paper}.  The differences between the
two algorithms will be outlined as part of the general explanation below.

The core of the \SLCT{} algorithm is quite simple: log lines are generally
created by substituting variable words into a fixed pattern, and analysis
of the frequency with which each word occurs can be used to determine
whether the word is variable or part of the fixed pattern.  This
classification can be used to group similar log lines and generate a
regex to match each group of log lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the input log lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with regex keywords
        that the parser will expand when it loads the rule (see the
        regex entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        regexes; it replaces a large number of variable words with fixed
        words, improving the subsequent classification of words as fixed or
        variable.  The new log lines are written to a temporary file, which
        all subsequent stages use instead of the original input file.

@@

            \item [pid] The \pid{} of the program that logged the line.

@@

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
        client hostname and \IP{} address.

@@

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an \SMTP{}
        command from the remote client: log the rejection with a mail if
        there is a queueid in the log line, or with the connection if not.

@@

    \item Mail enters the system via \SMTP{} or local submission.

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \SMTP{}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: PICKUP\@; flowchart:~2).

\SMTP{} is more complicated:

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client \IP{} address or hostname, HELO hostname, etc.\
        (action: REJECTION\@; flowchart:~4); acceptance of one or more
        mails (action: CLONE\@; flowchart:~5); or some interleaving of
        both.

    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
        Postfix has rejected any \SMTP{} commands the data will be saved to
        the database; if not there will not be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

    \item If one or more mails were accepted there will be more log entries
        for those mails later, see \sectionref{mail-delivery}.

@@

The mail lacks a queueid until it has been accepted, so log lines must
first be correlated by the \daemon{smtpd} \pid{}, then transition to being
correlated by the queueid.  This is relatively minor, but does require:

@@

    \item Most importantly: every section of code must know whether it
        needs to lookup the data structures by \pid{} or queueid.

@@

Multiple independent mails may be delivered across one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once the
mail has been accepted its log entries will not be correlated by \pid{} any
more, its queueid will be used instead (except when timeouts occur during
the data phase \sectionref{timeouts-during-data-phase}).  If the original
connection has any useful data (e.g.\ rejections) it will be saved to the
database when the client disconnects.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

    \item The client issues the RSET command to reset the \SMTP{} session.

@@

    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
        BD07D3C49

@@

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
        delivery is performed by the \SMTP{} client; local delivery means
        local to the server, i.e.\ an address the server is final
        destination for.} In this case the mail may already have been
        created (described above) and the unknown origin flag will be
        cleared; if not a new data structure will be created.  In both
        cases the re-injected mail is marked as a child of the original
        mail.  The log line in question is:

@@

    \item If the mail is of unknown origin it is assumed to be a child mail
        whose parent has not yet been identified (action: COMMIT\@;
        flowchart:~15).  Mark the mail as ready for entry in the database
        (flowchart:~16), and wait for the parent to deal with it
        (flowchart:~17).  The mail should not have subsequent log entries;
        only its parent should refer to it.

@@

    \item The sender address is \verb!<>!.

    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
        messages associated with the mail, indicating it was generated
        internally by Postfix, not accepted via \SMTP{} or submitted
        locally by \daemon{postdrop}.

@@

    \item The sender address must be \verb!<>!.

@@

Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient which
it continues with until delivery is complete.  Microsoft Outlook is one
client that behaves in this fashion; other clients may act in a similar
way.  An example dialogue exhibiting this behaviour is presented below
(lines starting with a three digit number are sent by the server, the other
lines are sent by the client):

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is discarded.  There will be no further
entries logged for such mails, so without identifying and discarding them
they accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \IP{} address, but the database schema
requires many more fields be populated (see \sectionref{connections table}
and \sectionref{results table}).  This heuristic is quite restrictive, and
it appears there is little scope for false positives; if there are any
false positives there will be warnings when the next log line for that mail
is parsed.  False negatives are less likely to be detected: there may be
queueid clashes (and warnings) if mails remain in the state tables after
they should have been removed, otherwise the only way to detect false
negatives is to examine the state tables after each parsing run.

@@

    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
        which passes through Postfix will have a \daemon{cleanup} line;
        lack of a \daemon{cleanup} line is a sure sign the mail did not
        make it too far.

@@

The DATA phase of the \SMTP{} conversation is where the headers and body of
the mail are transferred.  Sometimes there is a timeout or the connection
is lost\footnote{For brevity's sake \textit{timeout\/} will be used
throughout this section, but everything applies equally to lost
connections.} during the DATA phase; when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded with the connection data instead, which is saved.

@@

The list of valid combinations is explained below.  Every mail will
additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
both.

@@

The aborted delivery attempts described in
\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
delivery attempts described in this section only occur four times in the
\numberOFlogFILES{} log files used for testing.  The symptoms are the same
as in \sectionref{aborted-delivery-attempts}, except that there
\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
anything in the log file to explain why there are no further log messages.
The only way to detect these mails is to periodically scan all mails in the
state tables, deleting any mails displaying the following characteristics:

@@

    \item IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LINES\@?  There
        must be exactly two \daemon{smtpd} and one \daemon{cleanup} log
        entries for the mail, with no additional log entries.

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted the parser cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail (because it is the result of a mail loop, mail bomb, or
joe~job) before Postfix has had a chance to try to deliver it.  In this
case the recipient address will not have been logged, so a dummy recipient
address needs to be added, as every mail is required by the database schema
(\sectionref{results table}) to have at least one recipient.  This
complication has been observed in XXX/\numberOFlogFILES{} log files, but
typically when it does occur there will be many instances of it, closely
grouped.

@@

    \item Each new Postfix release requires new rules to be written or
        existing rules modified to cope with the new or changed log lines.
        Similarly using a new \DNSBL{}, a new policy server, or new
        administrator-defined rejection messages require new rules.

>>>>

    \item The log files are whole and complete: nothing has been removed,
        either deliberately or accidentally (e.g.\ log file rotation gone
        awry, file system filling up, logging system unable to cope with
        the volume of log messages).  On a well run system it is extremely
        unlikely that any of these problems will arise, though it is of
        course possible, particularly when suffering a deluge of spam or a
        mail loop.

@@

%    \item [cluster\_group] A reference to the \texttt{cluster\_group}
%        table.  That table is used by the Decision Tree algorithm
%        described in a separate document.

@@

    \item [server\_ip] The \gls{IP} address (IPv4 or IPv6) of the server:
        the local address when receiving mail, the remote address when
        sending mail.

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the \gls{IP} address could not be resolved to a
        hostname via DNS\@.

    \item [client\_ip] The client \gls{IP} address (IPv4 or IPv6): the
        remote address when receiving mail, the local address when sending
        mail.

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the \gls{IP} address could not be resolved to a
        hostname via DNS\@.

    \item [helo] The hostname used in the HELO command.  The HELO hostname
        occasionally changes during a connection, presumably because spam
        or virus senders think it is a good idea.  By default Postfix only
        logs the HELO hostname when it rejects an \gls{SMTP} command, but
        it is quite simple to rectify this, as described in
        \sectionref{logging helo}.

@@

    \item [start] The timestamp of the first log line, in seconds since the
        epoch.\glsadd{Epoch}

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \gls{SMTP} command ---
        a mechanism that is quite useful for testing new restrictions.
        This field will be 1 if the log line parsed was a warning rather
        than a real rejection, or 0 for a real rejection.  This field
        should be ignored if the result is not a rejection, i.e.\ the
        action field of the associated rule is not \texttt{REJECTION}.

    \item [smtp\_code] The \gls{SMTP} code associated with the log line.
        In general an \gls{SMTP} code is only present in a log line for a
        rejection or final delivery; results initially missing an
        \gls{SMTP} code will duplicate the \gls{SMTP} code of other results
        in that connection.  Some final delivery log lines do not contain
        an \gls{SMTP} code: in those cases the code is specified by the
        rule, based on the success or failure represented by the log line.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from a \gls{DNSBL}\@.

    \item [timestamp] The time at which the log line was logged, in seconds
        since the epoch.\glsadd{Epoch}

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \gls{DNSBL}
services, policy servers, or custom rejection messages.  To facilitate
parsing new log lines, the parser's design separates parsing rules from
parsing actions: adding new actions can be difficult, but adding new rules
to parse new rejection messages is trivial, and occurs much more
frequently.  A program is provided to ease the process of creating new
rules from unparsed log lines, based on the algorithm developed by Risto
Vaarandi for his \gls{SLCT}~\cite{slct-paper}.  The differences between the
two algorithms will be outlined as part of the general explanation below.

The core of the \gls{SLCT} algorithm is quite simple: log lines are
generally created by substituting variable words into a fixed pattern, and
analysis of the frequency with which each word occurs can be used to
determine whether the word is variable or part of the fixed pattern.  This
classification can be used to group similar log lines and generate a regex
to match each group of log lines.

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the input log lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \gls{IP} addresses, the
        standard start of rejection messages, etc.) with regex keywords
        that the parser will expand when it loads the rule (see the
        regex entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        regexes; it replaces a large number of variable words with fixed
        words, improving the subsequent classification of words as fixed or
        variable.  The new log lines are written to a temporary file, which
        all subsequent stages use instead of the original input file.

@@

            \item [pid] The \gls{pid} of the program that logged the line.

@@

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by \daemon{smtpd} \gls{pid}) and save both the
        client hostname and \gls{IP} address.

@@

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an \gls{SMTP}
        command from the remote client: log the rejection with a mail if
        there is a queueid in the log line, or with the connection if not.

@@

    \item Mail enters the system via \gls{SMTP} or local submission.

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \gls{SMTP}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: PICKUP\@; flowchart:~2).

\gls{SMTP} is more complicated:

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client \gls{IP} address or hostname, HELO hostname,
        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
        both.

    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
        Postfix has rejected any \gls{SMTP} commands the data will be saved
        to the database; if not there will not be any data to save (any
        mails accepted will already have been cloned so their data is in
        another data structure).

    \item If one or more mails were accepted there will be more log lines
        for those mails later, see \sectionref{mail-delivery}.

@@

The mail lacks a queueid until it has been accepted, so log lines must
first be correlated by the \daemon{smtpd} \gls{pid}, then transition to
being correlated by the queueid.  This is relatively minor, but does
require:

@@

    \item Most importantly: every section of code must know whether it
        needs to lookup the data structures by \gls{pid} or queueid.

@@

Multiple independent mails may be delivered across one connection: this
requires the algorithm to clone the current data as soon as a mail is
accepted, so that subsequent mails will not trample over each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once the
mail has been accepted its log lines will not be correlated by \gls{pid}
any more, its queueid will be used instead (except when timeouts occur
during the data phase \sectionref{timeouts-during-data-phase}).  If the
original connection has any useful data (e.g.\ rejections) it will be saved
to the database when the client disconnects.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

    \item The client issues the RSET command to reset the \gls{SMTP}
        session.

@@

    \item [Delivered to a remote \gls{SMTP} server:] 250 Ok: queued as
        BD07D3C49

@@

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
        delivery is performed by the \gls{SMTP} client; local delivery
        means local to the server, i.e.\ an address the server is final
        destination for.} In this case the mail may already have been
        created (described above) and the unknown origin flag will be
        cleared; if not a new data structure will be created.  In both
        cases the re-injected mail is marked as a child of the original
        mail.  The log line in question is:

@@

    \item If the mail is of unknown origin it is assumed to be a child mail
        whose parent has not yet been identified (action: COMMIT\@;
        flowchart:~15).  Mark the mail as ready for entry in the database
        (flowchart:~16), and wait for the parent to deal with it
        (flowchart:~17).  The mail should not have subsequent log lines;
        only its parent should refer to it.

@@

    \item The sender address is \verb!<>!.\glsadd{<>}

    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
        messages associated with the mail, indicating it was generated
        internally by Postfix, not accepted via \gls{SMTP} or submitted
        locally by \daemon{postdrop}.

@@

    \item The sender address must be \verb!<>!.\glsadd{<>}

@@

Some mail clients behave unexpectedly during the \gls{SMTP} dialogue: the
client aborts the first delivery attempt after the first recipient is
accepted, then makes a second delivery attempt for the same recipient which
it continues with until delivery is complete.  Microsoft Outlook is one
client that behaves in this fashion; other clients may act in a similar
way.  An example dialogue exhibiting this behaviour is presented below
(lines starting with a three digit number are sent by the server, the other
lines are sent by the client):

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is discarded.  There will be no further
entries logged for such mails, so without identifying and discarding them
they accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \gls{IP} address, but the database
schema requires many more fields be populated (see \sectionref{connections
table} and \sectionref{results table}).  This heuristic is quite
restrictive, and it appears there is little scope for false positives; if
there are any false positives there will be warnings when the next log line
for that mail is parsed.  False negatives are less likely to be detected:
there may be queueid clashes (and warnings) if mails remain in the state
tables after they should have been removed, otherwise the only way to
detect false negatives is to examine the state tables after each parsing
run.

@@

    \item Is the mail missing its \daemon{cleanup} log line?  Every mail
        which passes through Postfix will have a \daemon{cleanup} line;
        lack of a \daemon{cleanup} line is a sure sign the mail did not
        make it too far.

@@

The DATA phase of the \gls{SMTP} conversation is where the headers and body
of the mail are transferred.  Sometimes there is a timeout or the
connection is lost\footnote{For brevity's sake \textit{timeout\/} will be
used throughout this section, but everything applies equally to lost
connections.} during the DATA phase; when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded with the connection data instead, which is saved.

@@

The list of valid combinations is explained below.  Every mail will
additionally have log line from \daemon{cleanup} and \daemon{qmgr}; any
mail may also have log line from \daemon{bounce}, \daemon{postsuper}, or
both.

@@

The aborted delivery attempts described in
\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
delivery attempts described in this section only occur four times in the
\numberOFlogFILES{} log files used for testing.  The symptoms are the same
as in \sectionref{aborted-delivery-attempts}, except that there
\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
anything in the log file to explain why there are no further log lines.
The only way to detect these mails is to periodically scan all mails in the
state tables, deleting any mails displaying the following characteristics:

@@

    \item IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LINES\@?  There
        must be exactly two \daemon{smtpd} and one \daemon{cleanup} log
        lines for the mail, with no additional log lines.

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted the parser cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail (because it is the result of a mail loop\glsadd{mail
loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe job}) before
Postfix has had a chance to try to deliver it.  In this case the recipient
address will not have been logged, so a dummy recipient address needs to be
added, as every mail is required by the database schema
(\sectionref{results table}) to have at least one recipient.  This
complication has been observed in XXX/\numberOFlogFILES{} log files, but
typically when it does occur there will be many instances of it, closely
grouped.

@@

    \item Each new Postfix release requires new rules to be written or
        existing rules modified to cope with the new or changed log lines.
        Similarly using a new \gls{DNSBL}, a new policy server, or new
        administrator-defined rejection messages require new rules.

### 2008-11-22 doc/logparser-introduction.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the author's first real foray into processing
Postfix log files.}  Very few will have performed in-depth parsing and
analysis of their log files, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this kind of processing is the
unstructured nature of Postfix log files, where each log line was added on
an ad hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\DNSBL{}\footnote{This document is supplied with a glossary, see
\textsection\ref{Glossary}.} returns a different explanatory message;
policy servers may log different messages depending on the characteristics
of the connection.  There are many ways in which the log lines may differ
between servers, even within the same organisation: servers may be
configured differently, or running different versions of Postfix.  This
paper documents the difficult process of parsing Postfix log files,
presenting \PLP{}, a program that parses Postfix log files and places the
resulting data into a database for later use.  The gathered data can then
be used to optimise current anti-spam measures, provide a baseline to test
new anti-spam measures against, or to produce statistics showing how
effective those measures are.  Numerous other uses are possible for such
data: improving server performance by identifying troublesome destinations
and reconfiguring appropriately; identifying regular high volume uses
(e.g.\ customer newsletters) and restricting those uses to off-peak times;
detecting virus outbreaks that propagate via email; as a base for billing
customers on a shared server.  Preserving the raw data enables users to
develop a multitude of uses far beyond those conceived of by the author.

@@

Section~\ref{background} provides background information useful in
understanding the thesis, parser, and architecture.  It introduces the idea
of using a database schema as an \API{}, providing an interface to the data
gathered that is language-neutral.  The novel separation of rules, actions
and framework is discussed, giving the reasons that approach was taken when
designing the parser.

Section~\ref{state of the art review} reviews both the previously published
research in this area and other available Postfix log parsers, discussing
why they were deemed unsuitable for the task, including why they could not
be improved or expanded upon.

@@

The bibliography contains references to the resources used in developing
the algorithm, writing the program, and preparing this thesis.  Also
listed are some additional resources expected to be helpful in
understanding \SMTP{}, Postfix, anti-spam techniques, or the thesis.

Appendix~\ref{Glossary} provides a glossary of terms used in the thesis.

Appendix~\ref{Acronyms} provides a list of acronyms used in the thesis.

>>>>

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the author's first real foray into processing
Postfix log files.}  Very few will have performed in-depth parsing and
analysis of their log files, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this kind of processing is the
unstructured nature of Postfix log files, where each log line was added on
an ad hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\gls{DNSBL}\footnote{This document is supplied with a glossary, see
\textsection\ref{Glossary}.} returns a different explanatory message;
policy servers may log different messages depending on the characteristics
of the connection.  There are many ways in which the log lines may differ
between servers, even within the same organisation: servers may be
configured differently, or running different versions of Postfix.  This
paper documents the difficult process of parsing Postfix log files,
presenting \gls{PLP}, a program that parses Postfix log files and places
the resulting data into a database for later use.  The gathered data can
then be used to optimise current anti-spam measures, provide a baseline to
test new anti-spam measures against, or to produce statistics showing how
effective those measures are.  Numerous other uses are possible for such
data: improving server performance by identifying troublesome destinations
and reconfiguring appropriately; identifying regular high volume uses
(e.g.\ customer newsletters) and restricting those uses to off-peak times;
detecting virus outbreaks that propagate via email; as a base for billing
customers on a shared server.  Preserving the raw data enables users to
develop a multitude of uses far beyond those conceived of by the author.

@@

Section~\ref{background} provides background information useful in
understanding the thesis, parser, and architecture.  It introduces the idea
of using a database schema as an \gls{API}, providing an interface to the
data gathered that is language-neutral.  The novel separation of rules,
actions and framework is discussed, giving the reasons that approach was
taken when designing the parser.

Section~\ref{state of the art review} reviews both the previously published
research in this area and other available Postfix log file parsers,
discussing why they were deemed unsuitable for the task, including why they
could not be improved or expanded upon.

@@

The bibliography contains references to the resources used in developing
the algorithm, writing the program, and preparing this thesis.  Also
listed are some additional resources expected to be helpful in
understanding \gls{SMTP}, Postfix, anti-spam techniques, or the thesis.

Appendix~XXX provides a glossary of terms used in the thesis.

Appendix~XXX provides a list of acronyms used in the thesis.
